Equivalence Hypergraphs: E-Graphs for Monoidal Theories

### 2024-06-23 slides/slides.tex
slides fixes
slides update

%!LW recipe=xelatex

\documentclass[aspectratio=169]{beamer}

\usepackage{tikz}
\usepackage{tikz-network}
\usepackage{tikz-cd}
\usepackage{tikzit}
\usepackage{subcaption}
\usepackage{minted}

@@

\usetheme{metropolis}
% \usetheme{awesome}

@@

% \setbeamertemplate{footline}{\insertshortauthor~(\insertshortinstitute)}
% \usepackage{custom_footer}

@@

\begin{frame}{Table of contents}
    \setbeamertemplate{section in toc}[sections numbered]
    \tableofcontents[hideallsubsections]
\end{frame}

\section{What and Why?}

\begin{frame}
    Traditionally, e-graphs are used for \alert{theories} over \alert{algebraic} signatures with a single sort of composition $()$

    \begin{example}[Commutative monoid]

@@

    However, there are more general theories that describe the interaction of processes, \textit{e.g.}, quantum
\end{frame}

@@

\begin{example}
    \[
    (\mu\;;\sigma_{1,1}\;;i \otimes id_{1}) \otimes (\mu\;;(id_{1}\otimes id_{1}))
    \]
\end{example}

@@

Such signatures also come with \alert{built-in} equations

@@

\vspace{-1.5em}

@@

    The equations above are the laws of a (strict) symmetric monoidal category
    \vfill

    One gets a semantics of an SMT by building a functor for the corresponding \textbf{PROP}, \textit{e.g.} a functor into a category of vector spaces \textbf{$\text{Vect}_{k}$}
\end{frame}

@@

@@

\begin{minted}{c++}
    enum CommutativeComonoids {
        "mu" = Mu,
        "i" = I,
        ";" = Compose([Id;2]),
        "tensor" = Tensor([Id;2]),
        "id" = Identity,
        "I" = Unit,
        "sym" = Sym    
    }
\end{minted}

@@

\begin{frame}
    However, the buit-in equations (the laws of monoidal category) lead to an explosion of the number of e-classes

    \begin{example}
        \vspace{1em}
        E-graph for
        \[
            (\mu\;;\sigma_{1,1}\;;i \otimes id_{1}) \otimes (\mu\;;(id_{1}\otimes id_{1}))
        \]

        \begin{figure}
            \includegraphics[width=\linewidth]{figures/dot_5.pdf}
        \end{figure}

        \end{example}

        \pause

        Can we somehow get rid of these equations?
\end{frame}

@@

\begin{frame}
So \alert{yes}, we can get rid of these equations if we use \alert{hypergraphs}
\vfill
The work next formalises e-graphs for SMTs that would use string diagrams (hypergraphs) as a representation for terms
\end{frame}

@@

\begin{frame}{}
    \vfill
    Now equivalence classes represent equivalent subgraphs rather than subtrees
    \vfill
    Such combinatorial structures are shown to be \textit{sound} and \textit{complete} in terms of representing terms of enriched SMTs
    \vfill
    String diagrammatic reasoning becomes e-hypergraph rewriting
    \vfill
    The latter is defined as a \textit{double-pushout rewriting} which is a convenient framework to study properties of a rewriting system, e.g. confluence applicability of parallel rewrites etc.
\end{frame}

@@

@@

    We can show that our construction is indeed a generalisation of e-graphs

    The latter gives a categorical presentation of e-graphs and provides a framework for reasoning about them in terms of e-hypergraph rewriting
\end{frame}

@@

    \[
    \scalebox{0.5}{
    \tikzfig{figures/e-graphs-translation-3}
    }
    \]
    Can be rendered as
    \[
        % \hspace{1.3cm}
        \scalebox{0.38}{
        \tikzfig{../figures/categorical-semantics/egraph-translation-step-by-step-a-b}
        }
    \]
\end{frame}

\begin{frame}{}
    \vfill
    Lambda calculus is a model for a Cartesian Closed SMT and the benefits of using string diagrams for lambda calculus is that they are automatically quotiented by $\alpha$-equivalence (variables just become wires) and by linear substitution
    \vfill
    Thus we can represent e-graphs for lambda calculus with string diagrams for an enriched Cartesian Closed SMT
\end{frame}

\begin{frame}
    \[
        % \hspace{1.3cm}
        \scalebox{0.4}{
        \tikzfig{../figures/categorical-semantics/e-graph-substitution}
        }
    \]
    \[
        % \hspace{1.3cm}
        \scalebox{0.38}{
        \tikzfig{../figures/categorical-semantics/e-string-substitution-2}
        }
    \]
\end{frame}

>>>>

%!LW recipe=xelatex
\documentclass[aspectratio=169]{beamer}

\usepackage{tikz}

\usepackage{tikz-network}
\usepackage{tikz-cd}
\usepackage{tikzit}
\usepackage{subcaption}
\usepackage{minted}
\usepackage{mathtools}

% \hypersetup{
%     colorlinks=true,
%     linkcolor=gray,
%     filecolor=magenta,      
%     urlcolor=cyan,
%     pdftitle={Overleaf Example},
%     pdfpagemode=FullScreen,
%     }

@@

\usetheme[subsectionpage=simple]{metropolis}
% \usetheme{awesome}

@@

\makeatletter
\metropolis@disablesubsectionpage
\makeatother

\newcommand{\bsubsection}[1]{\subsection{$\bullet$ #1}}

@@

\begin{frame}{Table of contents}
    \small
    \setbeamertemplate{section in toc}[sections numbered]
    \tableofcontents
\end{frame}

\section{What and Why?}
\bsubsection{Algebraic theories}

\begin{frame}
    Traditionally, e-graphs are used for \alert{theories} over \alert{algebraic} signatures with a single sort of composition $()$
    \pause
    \vfill
    \begin{example}[Commutative monoid]

@@

    % However, there are more general theories that describe the interaction of processes, \textit{e.g.}, quantum
\end{frame}

\bsubsection{Monoidal theories}

@@

\pause
\begin{example}
    \[
    (\mu\;;(\sigma_{1,1}\;;i \otimes id_{1})) \otimes (\mu\;;(id_{1}\otimes id_{1}))
    \]
\end{example}

@@

Such signatures also come with \alert{built-in} equations (the laws of a symmetric monoidal category)

@@

\begin{frame}

@@

    One gets a semantics of an SMT by building a functor for the corresponding \textbf{PROP}, \textit{e.g.} a functor into a category of vector spaces \textbf{$\text{Vect}_{k}$}

    If the functor has nice properties, we can reason about the target category (e.g. \textbf{$\text{Vect}_{k}$}) syntactically
\end{frame}

@@

\bsubsection{Why e-graphs struggle with monoidal theories}

@@

\begin{minted}{c++}
    enum CommutativeComonoids {
        "mu" = Mu,
        "i" = I,
        ";" = Compose([Id;2]),
        "parallel" = Parallel([Id;2]),
        "id" = Identity,
        "I" = Unit,
        "sym" = Sym    
    }
\end{minted}

@@

\begin{frame}
    However, the laws of monoidal category lead to an explosion of the number of e-classes
\end{frame}

\begin{frame}
    \begin{example}
        \vspace{1em}
        E-graph for
        \[
            (\mu\;;(\sigma_{1,1}\;;i \otimes id_{1})) \otimes (\mu\;;(id_{1}\otimes id_{1}))
        \]

        Before saturation
        \vspace{-2em}
        \begin{figure}
            \includegraphics[scale=0.4]{figures/egraph_before_saturation.pdf}
        \end{figure}
    \end{example}
\end{frame}

\begin{frame}
    \begin{example}
        \vspace{1em}
        E-graph for
        \[
            (\mu\;;(\sigma_{1,1}\;;i \otimes id_{1})) \otimes (\mu\;;(id_{1}\otimes id_{1}))
        \]
        After saturation
        \begin{figure}
            \includegraphics[width=\linewidth]{figures/dot_5.pdf}
        \end{figure}

        \end{example}
\end{frame}

\begin{frame}[standout]
    Can we somehow get rid of these equations?
\end{frame}

\bsubsection{String diagrams and hypergraphs}

@@

\begin{frame}
    \begin{example}
        \begin{minipage}{0.45\linewidth}
        Assuming $\mu \coloneqq $\raisebox{-1.5em}{\scalebox{0.6}{\begin{tikzpicture}
            \begin{pgfonlayer}{nodelayer}
                \node [style=round box] (0) at (0.5, 0.5) {$\mu$};
                \node [style=none] (1) at (0.5, 1.5) {};
                \node [style=none] (2) at (0, -0.5) {};
                \node [style=none] (3) at (1, -0.5) {};
                \node [style=none] (4) at (0.25, 0.25) {};
                \node [style=none] (5) at (0.75, 0.25) {};
            \end{pgfonlayer}
            \begin{pgfonlayer}{edgelayer}
                \draw (0) to (1.center);
                \draw [bend right=15] (4.center) to (2.center);
                \draw [bend left=15] (5.center) to (3.center);
            \end{pgfonlayer}
        \end{tikzpicture}}}
        and $i \coloneqq \;\;$\raisebox{-0.2em}{\scalebox{0.6}{\begin{tikzpicture}
            \begin{pgfonlayer}{nodelayer}
                \node [style=none] (0) at (3.5, 1) {};
                \node [style=vertex] (1) at (3.5, 0.5) {};
            \end{pgfonlayer}
            \begin{pgfonlayer}{edgelayer}
                \draw (0.center) to (1);
            \end{pgfonlayer}
        \end{tikzpicture}        
        }} a string diagrammatic version of
        \[ 
            (\mu\;;(\sigma_{1,1}\;;i \otimes id_{1})) \otimes (\mu\;;(id_{1}\otimes id_{1}))
        \] looks like
    \end{minipage}
        \hfill
    \begin{minipage}{0.45\linewidth}
        \begin{tikzpicture}
                \begin{pgfonlayer}{nodelayer}
                    \node [style=round box] (0) at (0.5, 0.5) {$\mu$};
                    \node [style=none] (1) at (0.5, 1.5) {};
                    \node [style=none] (2) at (0, -0.25) {};
                    \node [style=none] (3) at (1, -0.25) {};
                    \node [style=none] (4) at (0.25, 0.25) {};
                    \node [style=none] (5) at (0.75, 0.25) {};
                    \node [style=vertex] (6) at (1, -2.25) {};
                    \node [style=none] (7) at (1, -1.75) {};
                    \node [style=round box] (8) at (3.5, 0.5) {$\mu$};
                    \node [style=none] (9) at (3.5, 1.5) {};
                    \node [style=none] (10) at (3, -0.25) {};
                    \node [style=none] (11) at (4, -0.25) {};
                    \node [style=none] (12) at (3.25, 0.25) {};
                    \node [style=none] (13) at (3.75, 0.25) {};
                    \node [style=none] (16) at (0.5, -1) {};
                    \node [style=none] (17) at (0, -1.75) {};
                    \node [style=none] (18) at (1, -1.75) {};
                    \node [style=none] (19) at (0, -2.25) {};
                    \node [style=none] (20) at (3, -2.25) {};
                    \node [style=none] (21) at (4, -2.25) {};
                \end{pgfonlayer}
                \begin{pgfonlayer}{edgelayer}
                    \draw (0) to (1.center);
                    \draw [bend right=15] (4.center) to (2.center);
                    \draw [bend left=15] (5.center) to (3.center);
                    \draw (6) to (7.center);
                    \draw (8) to (9.center);
                    \draw [bend right=15] (12.center) to (10.center);
                    \draw [bend left=15] (13.center) to (11.center);
                    \draw [in=30, out=-90] (3.center) to (16.center);
                    \draw [in=90, out=-135] (16.center) to (17.center);
                    \draw [in=315, out=90] (18.center) to (16.center);
                    \draw [in=-90, out=150] (16.center) to (2.center);
                    \draw (17.center) to (19.center);
                    \draw (10.center) to (20.center);
                    \draw (11.center) to (21.center);
                \end{pgfonlayer}
            \end{tikzpicture}
    \end{minipage}
    \vfill
    Which is already saturated with respect to the built-in equations
    \end{example} 
\end{frame}

\begin{frame}[standout]
So \alert{yes}, we can get rid of these equations if we use \alert{hypergraphs} (derived from string diagrams)
\end{frame}

\begin{frame}
\vfill
The work next formalises e-graphs for SMTs that would use string diagrams (hypergraphs) as a representation for terms in a bottom-up manner deriving this formalisation from an SMT itself
\vfill
\end{frame}

@@

\begin{frame}{}
    \vfill
    Now equivalence classes represent equivalent subgraphs rather than subtrees
    \pause
    \vfill
    \pause
    Such combinatorial structures are shown to be \textit{sound} and \textit{complete} for representing terms of enriched SMTs
    \vfill
    \pause
    String diagrammatic reasoning becomes e-hypergraph rewriting
    \vfill
    \pause
    The latter is defined as a \textit{double-pushout rewriting} which is a convenient framework to study properties of a rewriting system, e.g. confluence, applicability of parallel rewrites etc.
    \vfill
    \pause
    Note that they absorb the laws of symmetric monoidal category but not the structural rules for $+$
\end{frame}

@@

\bsubsection{(acyclic) E-graphs are e-hypergraphs for Cartesian monoidal theories}

@@

    We can show that our construction is indeed a generalisation of e-graphs
\end{frame}

@@

    \[
    \scalebox{0.5}{
    \tikzfig{figures/e-graphs-translation-3}
    }
    \]\footnote{[1]Willsey, M., Nandi, C., Remy Wang, Y., Flatt, O., Tatlock, Z., and Panchekha, P., “egg: Fast and Extensible Equality Saturation”, 2020.}
    Can be rendered as
    \[
        % \hspace{1.3cm}
        \scalebox{0.38}{
        \tikzfig{../figures/categorical-semantics/egraph-translation-step-by-step-a-b}
        }
    \]
\end{frame}

\begin{frame}
    \vfill
    This serves as a sanity check for our construction, provides a categorical presentation of e-graphs and therefore a possible framework for reasoning about them in terms of e-hypergraph rewriting
    \vfill
\end{frame}

\bsubsection{Support for binding}

\begin{frame}{}
    \small
    \vfill
    Lambda calculus is a model for a Cartesian Closed SMT and the benefits of using string diagrams for lambda calculus is that they are automatically quotiented by $\alpha$-equivalence (variables just become wires) and by linear substitution
    \vfill
    Compare the following

    \begin{minipage}{0.3\linewidth}
        \[
        \scalebox{0.5}{
        \tikzfig{figures/e-graphs-binding-example-plain}
        }
        \]
    \end{minipage}
    \hfill
    \begin{minipage}{0.3\linewidth}
        \[
        \scalebox{0.5}{
        \tikzfig{figures/e-graphs-binding-example-indices}
        }
        \]
    \end{minipage}
    \hfill
    \begin{minipage}{0.3\linewidth}
        \[
        \scalebox{0.5}{
        \tikzfig{figures/e-graphs-binding-string}
        }
        \]
    \end{minipage}
    \vfill

\begin{frame}
    \vfill
    Thus we can represent e-graphs for lambda calculus with string diagrams for an enriched Cartesian Closed SMT
    \vfill
\end{frame}

\begin{frame}
    \[
        % \hspace{1.3cm}
        \scalebox{0.4}{
        \tikzfig{../figures/categorical-semantics/e-graph-substitution}
        }
    \]
    \tiny{[2]\href{https://pldi23.sigplan.org/details/egraphs-2023-papers/12/Optimizing-Beta-Reduction-in-E-Graphs}{Optimizing-Beta-Reduction-in-E-Graphs}}
    \[
        % \hspace{1.3cm}
        \scalebox{0.35}{
        \tikzfig{../figures/categorical-semantics/e-string-substitution-2}
        }
    \]
\end{frame}
